---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
# Load data from zip file
activity_data_raw <- read.csv(unz("activity.zip","activity.csv"))

# Convert the dates to date objects
activity_data_raw$date<- as.Date(activity_data_raw$date, format='%Y-%m-%d')

# It does not make sense to average over the missing valuse, but 
# It appears to be part of the exercise, to show that the valuse 
# change after dealing with the missing data. So I will use the raw data
# and deal with missing values when taking the mean.

#########################
# Remove Code To remove missing data
#activity_data <- activity_data_raw[!is.na(activity_data_raw$steps),]
# remove extra levels associated with missing data
#activity_data$date <- factor(activity_data$date)

```


## What is mean total number of steps taken per day?
```{r} 
total_steps <- by(activity_data_raw$steps, activity_data_raw$date, sum, na.rm=TRUE )
hist(total_steps, breaks=20)
mean_steps <- mean(total_steps)
median_steps <- median(total_steps)

```
The mean total number of steps is `r mean_steps`.    
The median total number of steps is `r median_steps`.  

## What is the average daily activity pattern?
```{r}
avg_step_pattern <- with(activity_data_raw, tapply(steps, interval, mean, na.rm=TRUE))
with(activity_data_raw, plot(1:288, avg_step_pattern, type="l", xlab = "Time Interval", ylab = "Average Steps"))
max_interval=as.numeric(which(avg_step_pattern==max(avg_step_pattern), arr.ind=TRUE))
```

The interval with the maximun number of steps on average across all days is `r max_interval`  

## Imputing missing values
```{r}
library(plyr) 
num_missing_vals = length(activity_data_raw[is.na(activity_data_raw$steps),1])
missing_vals<-activity_data_raw[is.na(activity_data_raw),]
foo <- by(missing_vals$steps, missing_vals$date, count)
# missing vals are complete days
# average day on each side to replace missing vals
dailyAve <- by(activity_data_raw$steps, activity_data_raw$date, mean)
dailyAvePad = as.numeric(c(0,0,0,dailyAve,0,0,0))
dailyAvePad[is.na(dailyAvePad)] <- 0
dailyAvePadFiltered<-filter (dailyAvePad, filter = c(1/2, 0, 1/2), sides=2)
dailyAveVec<-as.numeric(dailyAve)
dailyAveFill=dailyAvePadFiltered[4:62]
na_fill_vals <- dailyAveFill[is.na(dailyAve)]
dailyAveVec[is.na(dailyAveVec)] <- dailyAveFill[is.na(dailyAve)]
#generate list of dates to update
na_dates <- unique(activity_data_raw[is.na(activity_data_raw$steps),2])
filler_key <- data.frame(na_dates, na_fill_vals)
```

The number of missing values in the inital data set is `r num_missing_vals`.    

## Are there differences in activity patterns between weekdays and weekends?
```{r}
library(lattice)
library(reshape2)
library(lubridate)
tday <- function(x) {
  if (x<5) {
    return("weekday")
  } else {
    return("weekend")
  }
}

activity_data_raw$typeDay<-lapply(wday(activity_data_raw$date), tday)
activity_data_raw$typeDay<-factor(as.character(activity_data_raw$typeDay))
mdata <- melt(activity_data_raw, id=c("typeDay","date", "interval"), measure.vars = c("steps"))
goo <-dcast(mdata, typeDay+interval ~ variable, fun.aggregate=mean, na.rm=TRUE)
xyplot(goo$steps ~ goo$interval | goo$typeDay, type='l', layout=c(1,2), xlab = "Time Interval", ylab = "Average Steps")
```
